Towards a Machine Learning approach for Aggregate Computing

### 2021-11-05 paper-2021-woa-ac-ml.tex
expand integration perspective, more on reinforcement learning integration

\usepackage{tcolorbox}
\usepackage[colorinlistoftodos,prependcaption]{todonotes}
\usepackage{xcolor}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{minted}
\usepackage{subcaption}
\usepackage[hidelinks]{hyperref}
\usepackage[inline]{enumitem}
\usepackage[T1]{fontenc}
\usepackage{cleveref}
\usepackage{systeme}
\setminted[scala]{fontsize=\scriptsize,baselinestretch=1,bgcolor=white,frame=single}

@@

So, we claim that Machine Learning can help us in handling this "unseen" situation, leaving only the burden
 to developers of defining the right collective specifications.
%
This lead to an ``hybrid" approach where a part of behaviour is still expressed using common 
 Aggregate Computing abstractions and another part is distilled through learning~\cite{research}.
%
In this paper, as the first steps towards a so-called \hybridaggregate{}, we investigate how Aggregate Computing can be
 extended with Reinforcement Learning through a case
 study where we outline the performance improvement of the "hop count" program.
\todos{Outline}

@@

Aggregate Computing is studied in various simulated scenario (thanks to Alchemist, a meta-simulator for pervasive systems), rangin from smart cities to crodws, thaks to programming language and toolchain that support 
 this paradigm, such as ScaFi~\cite{DBLP:conf/ecoop/CasadeiV16}, FCCP~\cite{DBLP:conf/acsos/Audrito20} and Protelis~\cite{PianiniSAC2015}.

In the next sections, we show the field calculus operators that are essential to understand the case study presented and then
 we underline how each node computes in order to follow the aggregate program speficiation 
\subsection{Field Calculus operators}

\subsection{ScaFi DSL}

\subsection{Computational Model}
An aggregate program execution consists of a continuously node-point execution of a \round{}, 
 that is the atomical part of aggregate computing.
%
Each node can interact only with-in its neighbourhoods sharing the \export{}, 
 namely the result produced by the evaluation of a \round{}.

The main round steps consist in:
\begin{enumerate}
  \item execution context creation: where node collects information from the sensors, the neighbourhoods exports and the old
  program output
  \item program evaluation: the system-wide aggregate program is evaluated against the local context created by the nodes. 
  This produces the \export{}.
  \item export sharing: the export created will be shared within the current neighbourhoods perceived by the nodes
  \item actuations: giving the export produced, nodes may performing some actuation (e.g. motor activations, steering, ...)
\end{enumerate}

The rounds execution is completely asynchronous, so do not exist a global clock to coordinate the aggregate. 
 This makes it possible to scale easily with the node number size.

@@

\paragraph{Multi- and many-agent system}
%
An aggregate system is a multi-agent system
 and, often, a \emph{many}-agent system
 where a large number (hundreds or more)
 of autonomous entities are programmed to achieve 
 some collective behaviour by means of \emph{repeated} 
 sensing, computation, communication, and actuation steps.
%
Due to the high stochasticity of the environment,
 it is almost impossible to know and
 program the optimal behaviour for all agents in advance.
 This results in the need to create intelligent agents
 so that they can learn the optimal behaviour and adapt to environmental changes.
%
In recent decades there has been an emerging trend in the use of Reinforcement Learning
 in multi-agent settings, as a powerful, robust and adaptive learning paradigm.
 Progress has been considerable and a wide range of algorithms are now available.
% to expand and references
In general, there is not one technical solution in the multi-agent context.
 There are several (even orthogonal) viewpoints on which researchers have been focused.
 One line of research (that of game theory) has tried to find techniques that achieve equilibrium such as Nash-Q learning~\cite{nash-q}.
 Others exploiting coordination between agents. 
 Others have applied single RL (such as Hysteretic Q-learning~\cite{hysteretic-q}) to each agent and achieved good results (even if no formal proof exist).
%
 The main problem with most of the solutions available in the literature is that they consider a small number of agents (or at least test them on small games).

\paragraph{Neighbour-based or indirect (environment-mediated) interaction}
%
In aggregate systems, a device can only directly interact with neighbours.
%
Data flows may be implemented
 to support indirect communication across multiple hops;
 however, information from devices far away tends to be obsolete.
%
The environment can also be used, via stigmergy,
 for indirect communication;
 however, a device can typically only access 
  to a very local portion of the environment.
%
In other words, the system state is only \emph{partially observable}.
%
\suggestions{What kind of ML algorithms can be used here?}
\paragraph{Learning goal: collective behaviour}
%
The devices of an aggregate 
 must cooperatively learn the ``aggregate program'',
 namely the collective behaviour 
 that achieves a particular \emph{global goal}
 in a decentralised, resilient way.
%

\paragraph{Self-organisation, swarm robotics}
%
\suggestions{You can talk about similar problem structure (i.e. the collective follow the same local behaviour)}

@@

@@

\paragraph{Transient phase}
%
When an aggregate behaviour is carried out,
 we often focus on the stabilised outcomes,
 tolerating outcomes that are being generated
 during the transient phase.
%
However, intermediate results are often important,
 and invariants should hold throughout the entire computation.

\paragraph{Dynamic topology}
%
If we admit \emph{mobility} and \emph{failure},
 then we must also consider 
 the possibility of changes in neighbourhoods
 and hence the need of dealing with dynamic topologies.

\subsection{Integration perspectives}
Here we outline the different perspectives about \hybridaggregate{} with the aim
 of clarifying the current research gap and to help us in the definition of a research roadmap.
%
\paragraph{Application} Learning can be useful at different levels in our framework.
 Surely, what for us is most important is to improve our building blocks definition.
% 
Indeed, most of our results and applications are based on them.
%
The building block improvement may consist either of learning a correction factor or by learning the behaviour entirely.

On the other side, learning can be useful also for framework related stuff,
 mainly aim at "non-functional" aspects. 
 Trust mechanisms, evaluation frequency, energy consumption manager, package storage are all the things that currently need to be handle by hand.

\paragraph{Learning problem} 
Multi-agent learning can be divided in \textit{Concurrent} and \textit{Team} learning. 
In the foster, each agent perform the learning process and in the latter,
 the learning process is performed only by one agent and then shared with the collective.
% 
Team learning is usually used in the ``swarm'' system, due to the homogeneous behaviour 
 --- i.e. each agent performs the same program. 
 Here, a common reward signal exists and evaluate the overall collective behaviour.
 Aggregate Computing is quite near to this kind of setting.
% 
Indeed, each agent follows the \textit{same} aggregate program shared within the entire system.

On the other hand, Concurrent Learning, in \cite{csas-and-marl} seems to be usually applied in Collective
 (Self-)Adaptive System, so again a field near to Aggregate Computing. 
 In particular, Indipendedet Reinforcement Learning seems to reach good results in this kind of system, 
 even if no theorems exist as in the single-agent context.
%
With Concurrent learning though is still possible to reach a global collective behaviour convergence. 
 Indeed using a common reward signal, the collective may reach the same behaviour as in the team learning settings as done here~\cite{iima2008swarm}.

@@

Reinforcement Learning is predominantly used in CAS, 
 due to its flexibility and thanks to the plethora of approaches used in this scenario.
% 
In our case, Reinforcement Learning can be used both at the building block level and the framework level.
%
For the foster, the simplest scenario can be that agents learn a correction factor to improve a long-term reward signal.

\paragraph{Learning time}
Finally, learning can be performed \textit{offline}, \texttt{online} or \textit{offine an then online} (\textit{mixed time}).
% 
The foster is the easiest scenario.
 Here indeed, we can leverage simulation to know the right computational field.
% 
Supervised, Reinforcement Learning and even Evolutionary computing can be used here.
%
Mixed-time learning can be a mid-level complexity direction instead.
 Here, simulation can always be leveraged to reach a good behaviour, but at runtime, the behaviour can be adapted in front of new environmental situations. 
%
Reinforcement Learning is the most suitable mode here --- also the "meta" learning can help.
%
Finally, we can imagine that learning can be performed online. 
 It is a common situation where we do not know the environment and also simulation can be very hard to accomplish. 
 Here, again, Reinforcement learning seems to be the most suitable approach.

\section{Aggregate Computing with Reinforcement}
In this section, we describe our first result in Aggregate Computing and Machine learning combination.
% 
Hence we briefly describe how Reinforcement Learning can be injected in the computation loop,
 and what kind of Reinforcement Learning can be used with Aggregate Computing.

@@

For our first tests, 
 we decide to set up a Concurrent learning technique, by which each agent try to improve our current local policy.
%
In these settings, the environment became non-stationary making the learning process unstable.
 But, as we observe in collective self-adaptive systems, practically these solutions led to good results.

>>>>

\usepackage{tcolorbox}
\usepackage[colorinlistoftodos,prependcaption]{todonotes}
\usepackage{xcolor}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{minted}
\usepackage{subcaption}
\usepackage[hidelinks]{hyperref}
\usepackage[inline]{enumitem}
\usepackage[T1]{fontenc}
\usepackage{cleveref}
\usepackage{systeme}
\setminted[scala]{baselinestretch=1,bgcolor=white,frame=single}

@@

So, we claim that Machine Learning can help us in handling this "unseen" situation, leaving only the burden
 to developers of defining the right collective specifications.
%
This lead to an ``hybrid" approach where a part of behaviour is still expressed using common 
 Aggregate Computing abstractions and another part is distilled through learning~\cite{research}.
%
In this paper, as the first steps towards a so-called \hybridaggregate{},
 we investigate how Aggregate Computing can be extended with Reinforcement Learning through a case study,
 where we outline the performance improvement of the "hop count" program.
\todos{Outline}

@@

Aggregate Computing is studied in various simulated scenarios (thanks to Alchemist, a meta-simulator for pervasive systems), ranging from smart cities to crowds, thanks to programming language and toolchain that support 
 this paradigm, such as ScaFi~\cite{DBLP:conf/ecoop/CasadeiV16}, FCCP~\cite{DBLP:conf/acsos/Audrito20} and Protelis~\cite{PianiniSAC2015}.

In the next sections, we show the field calculus operators that are essential to understand the case study presented,
 and then we underline how each node computes in order to follow the aggregate program specification.
\subsection{Field Calculus operators}
The main constructs that captures the essential aspects for programming self-organising systems are:
\begin{itemize}
  \item \textit{Stateful field evolution} expression \mintinline[escapeinside=||]{scala}{|{\color{teal}rep}|(|$e_1$|) {|($x$) => $e_2$|}} describes a field evolving in time. 
  $e_1$ is the initial field value and the function $(x) => e_2$ define how the field changes at each execution.
  \item \textit{Neighbour interaction} expression \mintinline[escapeinside=||]{scala}{|\color{teal}nbr|{|$e$|}} a view
  of the field values in the surroundings of each device where neighbours are
  mapped to their evaluations of $e$. A series of \mintinline{scala}{*hood} operators leverage this map to produce relevant result. 
  For instance, $minHood$ return the minimum value 
  from the neighbourhoods.
  \item \textit{Domain partitioning} expression \mintinline[escapeinside=||]{scala}{|\color{teal}if|(|$e_0$|){|$e_1$|}{|$e_2$|}} splits the computational
  field into two non-communicating zones hosting isolated subcomputations:
  $e_1$ where $e_0$ is true, and $e_2$ where $e_0$ is false.
\end{itemize}
\subsection{Computational Model}
An aggregate program execution consists of a continuously node-point execution of a \round{}, 
 that is the atomical part of aggregate computing.
%
Each node can interact only with-in its neighbourhoods sharing the \export{}, 
 namely the result produced by the evaluation of a \round{}.

The main round steps consist in:
\begin{enumerate}
  \item execution context creation: where node collects information from the sensors, the neighbourhoods exports and the old
  program output
  \item program evaluation: the system-wide aggregate program is evaluated against the local context created by the nodes. 
  This produces the \export{}.
  \item export sharing: the export created will be shared within the current neighbourhoods perceived by the nodes
  \item actuations: giving the export produced, nodes may perform some actuation (e.g. motor activations, steering, ...)
\end{enumerate}

Rounds execution is completely asynchronous, so do not exist a global clock to coordinate the aggregate. 
 This makes it possible to scale easily with the node number size.

@@

\paragraph{Multi-and many-agent system}
%
An aggregate system is a multi-agent system
 and, often, a \emph{many}-agent system
 where a large number (hundreds or more)
 of autonomous entities are programmed to achieve 
 some collective behaviour by means of \emph{repeated} 
 sensing, computation, communication, and actuation steps.
%
Due to the high stochasticity of the environment,
 it is almost impossible to know and
 program the optimal behaviour for all agents in advance.
 This results in the need to create intelligent agents
 so that they can learn the optimal behaviour and adapt to environmental changes.
%
In recent decades there has been an emerging trend in the use of Reinforcement Learning
 in multi-agent settings, as a powerful, robust and adaptive learning paradigm.
 Progress has been considerable and a wide range of algorithms are now available.
% to expand and references
This field is a conjunction of Game theory and Reinforcement Learning, 
 and there are several (even orthogonal) viewpoints on which researchers have been focused.

First attempts from Reinforcement Learning viewpoint, 
 goes towards a so-called \textit{conccurent/independent learning}~\cite{DBLP:journals/tsmc/BusoniuBS08} approach, 
 where each agent learn locally against the whole environment~\cite{DBLP:conf/icml/Tan93}. 
Here various approaches have been proposed,
 such as Hysteretic Q-Learning (used to reduce non-stationarity)~\cite{hysteretic-q}, 
 Lenient Learning~\cite{DBLP:journals/jmlr/WeiL16}, and Frequency Maximum Q-value.
The pro of these approaches is that their complexity does not scale up with the number of agents, 
 but, unfortunately, the learning process is extremely non-stationary and non-stable.
 Furthermore, they are heuristic and does not exist any convergence proof to a stationary state (even if practically often convergence to a stable policy).

%
The game theory viewpoint instead has tried to find techniques that achieve equilibrium such as Nash-equilibrium of Pareto optimality.
 The first works towards this direction are Nash-Q-Learning~\cite{nash-q}, Friend-or-Foe Q-learning~\cite{DBLP:conf/icml/Littman01}, Minimax-Q.
The problem there is that is used with few agents and scale hardly with the application complexity.
%
So, current emerging trends tend to leverage a so-called centralised training and decentralised execution (CTDE) approach, by which 
 agents should leverage the system-wide knowledge at training time but at runtime, they act independently~\cite{DBLP:journals/aamas/Hernandez-LealK19}.
 Most promising approaches in this direction are MAPPO~\cite{DBLP:journals/corr/abs-2103-01955}, COMA~\cite{DBLP:journals/corr/FoersterFANW17} and MADDPG~\cite{DBLP:conf/nips/LoweWTHAM17}.

%
 The main problem with most of the solutions available in the literature is that they consider a small number of agents (or at least test them on small games).

\paragraph{Neighbour-based or indirect (environment-mediated) interaction}
%
In aggregate systems, a device can only directly interact with neighbours.
%
Data flows may be implemented
 to support indirect communication across multiple hops;
 however, information from devices far away tends to be obsolete.
%
The environment can also be used, via stigmergy,
 for indirect communication;
 however, a device can typically only access 
  to a very local portion of the environment.
%
In other words, the system state is only \emph{partially observable}.

In practice, CTDE technique aims to solve these problems, using a \emph{global} view at training time but
 a \emph{partial view} at execution time.

Furthermore, when the system node size is very large, novel works of Mean-Field Reinforcement Learning~\cite{DBLP:conf/ijcnn/ZhouZX21} are able to
 handle learning efficiently, reasoning about agent and agent neighbourhoods.
%
\paragraph{Learning goal: collective behaviour}
%
The devices of an aggregate 
 must cooperatively learn the ``aggregate program'',
 namely the collective behaviour 
 that achieves a particular \emph{global goal}
 in a decentralised, resilient way.
%

% \paragraph{Self-organisation, swarm robotics} it is condensed it other perspective, so I think it is better to avoid this paragraph.
%
% \suggestions{You can talk about similar problem structure (i.e. the collective follow the same local behaviour)}

@@

Typically, in CAS, single-agent RL brings to emergent behaviour and is used in various applications~\cite{DBLP:conf/icse/DAngeloGGGNPT19}.

@@

A way to handle this kind of rewards in multi-agent settings is \textit{difference rewards}~\cite{DBLP:conf/atal/DevlinYKT14}. It is used to capture agents contributions to the system's global performance.
\paragraph{Transient phase}
%
When an aggregate behaviour is carried out,
 we often focus on the stabilised outcomes,
 tolerating outcomes that are being generated
 during the transient phase.
%
However, intermediate results are often important,
 and invariants should hold throughout the entire computation.

\paragraph{Dynamic topology}
%
If we admit \emph{mobility} and \emph{failure} (rarely considered in related works, where node population is fixed),
 then we must also consider
 the possibility of changes in neighbourhoods
 and hence the need of dealing with dynamic topologies.

\subsection{Integration perspectives}
Here we outline the different perspectives about \hybridaggregate{} with the aim
 of clarifying the current research.
%
\paragraph{Application} Learning can be useful at different levels in our framework.
 Surely, what for us is most important is to improve our building blocks definition.
% 
Indeed, most of our results and applications are based on them.
%
The building block improvement may consist either of learning a correction factor or by learning the behaviour entirely.

On the other side, learning can be useful also for framework related stuff,
 mainly aim at "non-functional" aspects. 
 Trust mechanisms, evaluation frequency, energy consumption manager, package storage are all the things that could need to be handled by hand.

\paragraph{Learning problem} 
The learning problem in multi-agent settings can be pose as \textit{Independent} and \textit{Team} learning. 
In the foster, each agent perform the learning process and in the latter,
 the learning process is performed only by one agent and then shared with the collective.
% 
Team learning is usually used in the ``swarm'' system, due to the homogeneous behaviour 
 --- i.e. each agent performs the same program. 
 Here, a common reward signal exists and evaluate the overall collective behaviour.
 Aggregate Computing is quite near to this kind of setting.
% 
Indeed, each agent follows the \textit{same} aggregate program shared within the entire system.

On the other hand, Independent Learning is usually applied in Collective
 (Self-)Adaptive System \cite{csas-and-marl}, so again a field near to Aggregate Computing.
 In particular, Independent Reinforcement Learning seems to reach good results in this kind of system, 
 even if no theorems exist as in the single-agent context.
%
With Concurrent learning though is still possible to reach a global collective behaviour convergence. 
 Indeed using a common reward signal, the collective may reach the same behaviour \cite{iima2008swarm, nguyen2018swarm}.

@@

Reinforcement Learning is predominantly used in CAS, 
 due to its flexibility and thanks to the plethora of approaches used in this scenario~\cite{csas-and-marl}.
% 
In our case, Reinforcement Learning can be used both at the building block level and the framework level.
%
For the foster, the simplest scenario can be that agents learn a correction factor to improve a long-term reward signal.

\paragraph{Learning time}
Finally, learning can be performed \textit{offline}, \texttt{online} or \textit{offine an then online} (\textit{mixed time}).
% 
The foster is the easiest scenario.
 Here indeed, we can leverage simulation to know the right computational field and then use one of the novels CTDE techniques. 
% 
Supervised, Reinforcement Learning and even Evolutionary computing can be used here.
%
Mixed-time learning can be a mid-level complexity direction instead.
 Here, simulation can always be leveraged to reach a good behaviour, but at runtime, the behaviour can be adapted in front of new environmental situations. 
%
Finally, we can imagine that learning can be performed online. 
 It is a common situation where we do not know the environment and also simulation can be very hard to accomplish.

In mixed and online cases, Reinforcement Learning is the most suitable mode here --- also the "meta" learning can help.

\section{Aggregate Computing with Reinforcement}
In this section, we describe our first result in Aggregate Computing and Machine learning combination.
% 
Hence we briefly describe how Reinforcement Learning can be injected into the computation loop,
 and what kind of Reinforcement Learning can be used with Aggregate Computing.

@@

For our first tests, 
 we decide to set up an Independent learning technique, by which each agent try to improve our current local policy.
 Our approach is partially inspired by Swarm Q-Learning~\cite{nguyen2018swarm} and leverage Hysteretic Q-Learning~\cite{hysteretic-q} to 
 handle
 the non-stationarity in the environment.
%
In particular, each agent maintains a local Q table ($Q_i(s, a)$, where i is the agent id) following the standard Hysteretic update:
$$
Q(s, a) = 
$$
At the end of the episodes, each agent has collected its local experience, which can vary in different zones. So, we combine the N q table
 selecting the maximum value for each state-action pair:
$$
Q_{max}(s, a) = max_i(Q_i(s, a))
$$
Then, in the new episode, each agent starts with the $Q_max$ and try to refine it. 
 This max combination works because we assume a homogeneous behaviour, where each agent perform
 in the same way when reaching the same state.
